Foundations of community ecology

### 2025-06-18 foundations_of_community_ecology.Rmd
Initialise commentary point

>>>>

**IDEAS**

Do we have an issue because *only* a trait of j's should affect both i's contribution to the sum population size and the sum trait quantity omega, right? The existence of j shouldn't matter, unless we're multiplying 1 by j's trait. But we shouldn't double count j's existence and its trait value to predict how it affects i (or anything about i).

Or maybe the original formula isn't quite right. We need alpha to actually be imaginary. So maybe if we have 0 + i\alpha_{ij}, something like $(0 + i1)\alpha * (i\tau_{ij})$

Well, could try to have an individual with the property $(1 + 0.5i)$ and define ($\alpha = 1 + 0.1i$, but something seems off here. Then again, if we have fitness as 2, then set an individual and trait to $(1 + 0.5i)$, affected by competition of a j individual, $a_{ij} = (0 + 0.5i)$. I'm not sure. I can't get the density to change while also getting the trait to change. Unless it's something like $z = 1 + 0.5i$, then $a_{ij} = (0 + 0.1i)$, so $z + za$.

One last try before bed, if w = 2, let's say, $2 (1 + 0.5i) * (1 - (0 - 0i) * (1 + 0.4i))$, where $\alpha = (0 - 0i)$ and j is $(1 + 0.4i)$. But alpha is the effect of a *trait* on fitness, so something isn't quite right here.

Okay, try making a_{ij} into an imaginary number. This should work.

Deriving Price like
=====================

$$\frac{\Omega}{N} = \underbrace{\mathbb{E}[w z]}_{\text{Term 1}} + \underbrace{\mathbb{E}[w \Delta z]}_{\text{Term 2}} - \underbrace{\mathbb{E}\left[w \sum_{j=1}^N a_{ij} z_i^2 \right]}_{\text{Term 3}} - \underbrace{\mathbb{E}\left[w \sum_{j=1}^N \left( 2 a_{ij} z_i \Delta z_i + a_{ij} (\Delta z_i)^2 \right) \right]}_{\text{Term 4}}.$$

Then,

$$\bar{w} \bar{z}' = \underbrace{\text{Cov}(w, z) + \bar{w} \bar{z}}_{\text{Term 1}} + \underbrace{\mathbb{E}[w \Delta z]}_{\text{Term 2}} - \underbrace{\mathbb{E}\left[w \sum_{j=1}^N a_{ij} z_i^2 \right]}_{\text{Term 3}} - \underbrace{\mathbb{E}\left[w \sum_{j=1}^N \left( 2 a_{ij} z_i \Delta z_i + a_{ij} (\Delta z_i)^2 \right) \right]}_{\text{Term 4}}.$$

Then,

$$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - \mathbb{E}\left[w \sum_{j=1}^N a_{ij} z_i^2 \right] - \mathbb{E}\left[w \sum_{j=1}^N \left( 2 a_{ij} z_i \Delta z_i + a_{ij} (\Delta z_i)^2 \right) \right].$$
If we are willing to assume that all individuals $j$ have the same effect on a focal $i$, then we can rewrite as,

$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - N \mathbb{E}[w \alpha_i z_i^2] - N \mathbb{E}\left[w \alpha_i \left( 2 z_i \Delta z_i + (\Delta z_i)^2 \right) \right].$

This simplifies again,

$$\begin{align*}
\bar{w} \Delta \bar{z} &= \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - N \mathbb{E}\left[ w \alpha_i \left( z_i^2 + 2 z_i \Delta z_i + (\Delta z_i)^2 \right) \right] \\
&= \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - N \mathbb{E}\left[ w \alpha_i (z_i + \Delta z_i)^2 \right].
\end{align*}$$

$$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) + \mathbb{E}[w \Delta z] - N \mathbb{E}\left[ w \alpha_i (z_i + \Delta z_i)^2 \right].$$
Let's assume that $\Delta z_{i} = 0$, where the middle term will drop out and we can focus on population level effects on trait change (selection and drift) and the effect of interaction coefficients,

$$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) - N \mathbb{E}[w \alpha_i z_i^2].$$

Let's just drop the subscripts in the second term as in the first,

$$\bar{w} \Delta \bar{z} = \text{Cov}(w, z) - N \mathbb{E}[w \alpha z^2].$$

It's reassuring that the sign of z doesn't affect the second term. Verbally, 
I *think* the first term represents the change in a trait attributable to selection (and drift) in the absence of interspecific interactions, while the second term represents trait change attributable to selection attributable to interactions?

Note that the second term is the expected interaction between fitness and interaction strength $w \alpha$ times $z^2$, which embeds the variance of $z^{2}$ because,

$$E[z_{i}^{2}] = Var(z) + \bar{z}^{2}$$
